---
title: "MetaPhlAn4 Microbiome Analysis: HRSSL vs NN"
subtitle: "Alpha Diversity, Beta Diversity, and Differential Abundance (DESeq2 & edgeR)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
header-includes:
  - \usepackage{geometry}
  - \geometry{margin=1in}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "90%",
  cache      = FALSE
)
```

# Setup and Data Loading

## Load Libraries

```{r libraries}
# Bioconductor packages
library(phyloseq)
library(DESeq2)
library(edgeR)

# Ecological / stats
library(vegan)

# Plotting and tables
library(ggplot2)
library(kableExtra)
library(patchwork)
library(RColorBrewer)

# Tidyverse loaded last to avoid masking conflicts (dplyr vs S4Vectors)
library(tidyverse)
```

## Define File Paths

Update these paths to point to your actual data files.

```{r file-paths}
metaphlan_file <- "fitNNandHRSSLmerged.tsv"
metadata_file  <- "metadata_wsex.txt"
```

## Load MetaPhlAn4 Data

```{r load-metaphlan}
# Read the MetaPhlAn4 merged abundance table
mp_raw <- read.delim(metaphlan_file, header = TRUE, sep = "\t",
                     check.names = FALSE, stringsAsFactors = FALSE)
colnames(mp_raw)[1] <- "clade_name"

cat("Total rows:", nrow(mp_raw), "\n")
cat("Total samples:", ncol(mp_raw) - 1, "\n")
```

## Load Metadata

```{r load-metadata}
metadata <- read.delim(metadata_file, header = TRUE, sep = "\t",
                       stringsAsFactors = FALSE)
rownames(metadata) <- metadata$sample_id
metadata$source <- factor(metadata$source, levels = c("NN", "HRSSL"))

cat("Samples per group:\n")
print(table(metadata$source))
```

## Parse Taxonomy and Filter to Species Level

```{r parse-species}
# Extract species-level rows: contain "s__" but not "t__"
species_df <- mp_raw %>%
  dplyr::filter(grepl("s__", clade_name) & !grepl("t__", clade_name))

cat("Species-level taxa:", nrow(species_df), "\n")

# Parse the full taxonomy string into individual ranks
parse_taxonomy <- function(clade_names) {
  ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__")

  tax_mat <- matrix(NA, nrow = length(clade_names), ncol = length(ranks),
                    dimnames = list(NULL, ranks))

  for (i in seq_along(clade_names)) {
    parts <- unlist(strsplit(clade_names[i], "\\|"))
    for (j in seq_along(prefixes)) {
      match_part <- parts[grepl(paste0("^", prefixes[j]), parts)]
      if (length(match_part) > 0) {
        tax_mat[i, j] <- gsub(prefixes[j], "", match_part[1])
      }
    }
  }
  return(tax_mat)
}

tax_matrix <- parse_taxonomy(species_df$clade_name)

# Create clean species labels for row names
species_labels <- gsub(".*\\|s__", "", species_df$clade_name)
species_labels <- gsub("_", " ", species_labels)
rownames(tax_matrix) <- species_labels
```

## Build Abundance Matrix

```{r build-abundance-matrix}
# Extract numeric abundance columns, align with metadata
abund_mat <- as.matrix(species_df[, colnames(species_df) %in% metadata$sample_id])
rownames(abund_mat) <- species_labels

# Ensure sample order matches metadata
shared_samples <- intersect(colnames(abund_mat), rownames(metadata))
abund_mat <- abund_mat[, shared_samples]
metadata <- metadata[shared_samples, ]

cat("Pre-filter dataset:", nrow(abund_mat), "species x", ncol(abund_mat), "samples\n")
```

## Prevalence Filtering

A standard prevalence filter is applied: species must be present (non-zero abundance) in more than 10% of samples to be retained. This filter is applied once here, prior to normalisation, so that all downstream analyses (alpha diversity, beta diversity, differential abundance) operate on the same filtered taxon set.

```{r prevalence-filter}
prevalence_threshold <- 0.10  # >10% of samples

prevalence <- rowSums(abund_mat > 0) / ncol(abund_mat)
keep_taxa <- prevalence > prevalence_threshold

cat("Prevalence threshold: >", prevalence_threshold * 100, "% of samples (i.e., > ",
    round(prevalence_threshold * ncol(abund_mat), 2), " samples)\n", sep = "")
cat("Taxa retained: ", sum(keep_taxa), " of ", length(keep_taxa),
    " (", round(100 * sum(keep_taxa) / length(keep_taxa), 1), "%)\n", sep = "")

# Apply filter
abund_mat   <- abund_mat[keep_taxa, ]
tax_matrix  <- tax_matrix[keep_taxa, ]

cat("Post-filter dataset:", nrow(abund_mat), "species x", ncol(abund_mat), "samples\n")
```

## Build phyloseq Object

```{r build-phyloseq}
OTU <- otu_table(abund_mat, taxa_are_rows = TRUE)
TAX <- tax_table(tax_matrix)
SAM <- sample_data(metadata)

ps <- phyloseq(OTU, TAX, SAM)
ps
```

\newpage

# Normalisation

MetaPhlAn4 outputs relative abundances as percentages (summing to 100 per sample). We apply two standard normalisations used throughout the downstream analyses:

- **TSS (Total Sum Scaling):** converts percentages to proportions (0--1). Used for Bray-Curtis dissimilarity and alpha diversity, where the compositional scale is appropriate.
- **CLR (Centered Log-Ratio):** the standard transformation for compositional data analysis. Maps abundances into unconstrained Euclidean space, enabling distance metrics and ordinations that are not distorted by the compositional constraint. A pseudocount is added to zeros prior to log transformation.

```{r normalisation}
# --- TSS: convert percentages to proportions (sum to 1 per sample) ---
tss_mat <- sweep(abund_mat, 2, colSums(abund_mat), FUN = "/")

# Verify column sums equal 1
stopifnot(all(abs(colSums(tss_mat) - 1) < 1e-10))

cat("TSS normalisation applied — column sums:", unique(round(colSums(tss_mat), 4)), "\n")

# --- CLR: centered log-ratio transformation ---
# Add a pseudocount (minimum non-zero value per sample / 2) to handle zeros
clr_transform <- function(mat) {
  # mat: taxa (rows) x samples (cols), proportions
  clr_out <- matrix(NA, nrow = nrow(mat), ncol = ncol(mat),
                    dimnames = dimnames(mat))
  for (j in seq_len(ncol(mat))) {
    x <- mat[, j]
    # Per-sample pseudocount: half the smallest non-zero value
    pseudo <- min(x[x > 0]) / 2
    x_pseudo <- x + pseudo
    log_x <- log(x_pseudo)
    clr_out[, j] <- log_x - mean(log_x)
  }
  return(clr_out)
}

clr_mat <- clr_transform(tss_mat)

cat("CLR transformation applied — dimensions:", nrow(clr_mat), "taxa x", ncol(clr_mat), "samples\n")
```

\newpage

# Alpha Diversity

Alpha diversity is calculated from TSS-normalised proportions. We compute Shannon, Simpson, and Observed richness indices, then compare groups using Wilcoxon rank-sum tests.

```{r alpha-diversity}
# Calculate alpha diversity metrics on TSS-normalised data
alpha_df <- data.frame(
  sample_id = sample_names(ps),
  Shannon   = diversity(t(tss_mat), index = "shannon"),
  Simpson   = diversity(t(tss_mat), index = "simpson"),
  Observed  = colSums(tss_mat > 0),
  stringsAsFactors = FALSE
)
alpha_df <- dplyr::left_join(alpha_df, metadata, by = "sample_id")
```

## Statistical Tests

```{r alpha-stats}
alpha_metrics <- c("Shannon", "Simpson", "Observed")
alpha_tests <- lapply(alpha_metrics, function(metric) {
  wt <- wilcox.test(
    as.formula(paste(metric, "~ source")),
    data = alpha_df
  )
  data.frame(
    Metric  = metric,
    W       = wt$statistic,
    p.value = round(wt$p.value, 4),
    stringsAsFactors = FALSE
  )
})
alpha_test_df <- do.call(rbind, alpha_tests)
rownames(alpha_test_df) <- NULL

kable(alpha_test_df, booktabs = TRUE, longtable = TRUE,
      caption = "Wilcoxon rank-sum tests for alpha diversity (HRSSL vs NN)") %>%
  kable_styling(latex_options = c("repeat_header", "scale_down", "HOLD_position"))
```

## Alpha Diversity Plots

```{r alpha-plots, fig.height=4, fig.cap="Alpha diversity comparisons between HRSSL and NN groups."}
plot_alpha <- function(df, metric) {
  ggplot(df, aes(x = source, y = .data[[metric]], fill = source)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.15, size = 2, alpha = 0.6) +
    scale_fill_manual(values = c("NN" = "#4DAF4A", "HRSSL" = "#E41A1C")) +
    labs(y = metric, x = NULL) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
}

p_shan <- plot_alpha(alpha_df, "Shannon")
p_simp <- plot_alpha(alpha_df, "Simpson")
p_obs  <- plot_alpha(alpha_df, "Observed")

p_shan + p_simp + p_obs + plot_layout(ncol = 3)
```

\newpage

# Beta Diversity

Two complementary distance metrics are used:

- **Bray-Curtis dissimilarity** on TSS-normalised proportions --- the most widely used metric for microbiome beta diversity.
- **Aitchison distance** (Euclidean distance on CLR-transformed data) --- the compositionally appropriate metric that accounts for the constant-sum constraint inherent in relative abundance data.

## Bray-Curtis Dissimilarity (TSS) and PCoA

```{r beta-pcoa-bc, fig.height=5, fig.cap="PCoA of Bray-Curtis dissimilarity (TSS-normalised) coloured by group."}
# Transpose: samples in rows for vegan
tss_t <- t(tss_mat)

bc_dist <- vegdist(tss_t, method = "bray")
pcoa_bc <- cmdscale(bc_dist, k = 2, eig = TRUE)

# Variance explained
eig_pct_bc <- round(pcoa_bc$eig / sum(pcoa_bc$eig) * 100, 1)

pcoa_bc_df <- data.frame(
  Axis1 = pcoa_bc$points[, 1],
  Axis2 = pcoa_bc$points[, 2],
  source = metadata$source,
  sample_id = metadata$sample_id
)

ggplot(pcoa_bc_df, aes(x = Axis1, y = Axis2, colour = source)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", linewidth = 0.8, linetype = "dashed") +
  scale_colour_manual(values = c("NN" = "#4DAF4A", "HRSSL" = "#E41A1C")) +
  labs(
    title = "Bray-Curtis (TSS)",
    x = paste0("PCoA1 (", eig_pct_bc[1], "%)"),
    y = paste0("PCoA2 (", eig_pct_bc[2], "%)"),
    colour = "Group"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Aitchison Distance (CLR) and PCoA

```{r beta-pcoa-aitchison, fig.height=5, fig.cap="PCoA of Aitchison distance (CLR-transformed) coloured by group."}
clr_t <- t(clr_mat)

ait_dist <- vegdist(clr_t, method = "euclidean")
pcoa_ait <- cmdscale(ait_dist, k = 2, eig = TRUE)

eig_pct_ait <- round(pcoa_ait$eig / sum(pcoa_ait$eig) * 100, 1)

pcoa_ait_df <- data.frame(
  Axis1 = pcoa_ait$points[, 1],
  Axis2 = pcoa_ait$points[, 2],
  source = metadata$source,
  sample_id = metadata$sample_id
)

ggplot(pcoa_ait_df, aes(x = Axis1, y = Axis2, colour = source)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", linewidth = 0.8, linetype = "dashed") +
  scale_colour_manual(values = c("NN" = "#4DAF4A", "HRSSL" = "#E41A1C")) +
  labs(
    title = "Aitchison Distance (CLR)",
    x = paste0("PCoA1 (", eig_pct_ait[1], "%)"),
    y = paste0("PCoA2 (", eig_pct_ait[2], "%)"),
    colour = "Group"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## PERMANOVA

```{r permanova}
set.seed(42)
perm_bc  <- adonis2(bc_dist  ~ source, data = metadata, permutations = 999)
perm_ait <- adonis2(ait_dist ~ source, data = metadata, permutations = 999)

kable(as.data.frame(perm_bc), booktabs = TRUE, digits = 4,
      caption = "PERMANOVA — Bray-Curtis dissimilarity (TSS, 999 permutations)") %>%
  kable_styling(latex_options = c("HOLD_position"))

kable(as.data.frame(perm_ait), booktabs = TRUE, digits = 4,
      caption = "PERMANOVA — Aitchison distance (CLR, 999 permutations)") %>%
  kable_styling(latex_options = c("HOLD_position"))
```

## Test for Homogeneity of Dispersions (betadisper)

```{r betadisper}
# Bray-Curtis
bd_bc <- betadisper(bc_dist, metadata$source)
bd_bc_test <- permutest(bd_bc, permutations = 999)

cat("Betadisper — Bray-Curtis (TSS):\n")
cat("  F =", round(bd_bc_test$tab$F[1], 3), "\n")
cat("  p =", round(bd_bc_test$tab$`Pr(>F)`[1], 4), "\n\n")

# Aitchison
bd_ait <- betadisper(ait_dist, metadata$source)
bd_ait_test <- permutest(bd_ait, permutations = 999)

cat("Betadisper — Aitchison (CLR):\n")
cat("  F =", round(bd_ait_test$tab$F[1], 3), "\n")
cat("  p =", round(bd_ait_test$tab$`Pr(>F)`[1], 4), "\n")
```

\newpage

# Phylum-Level Composition

Relative abundance (%) at the phylum level from the original MetaPhlAn4 output. These values are displayed as reported (percentage scale) since stacked barplots represent proportional composition directly.

```{r phylum-barplot, fig.height=5, fig.cap="Relative abundance at the phylum level by sample, ordered by group."}
# Extract phylum-level rows from the original table
phylum_df <- mp_raw %>%
  dplyr::filter(grepl("p__", clade_name) & !grepl("c__", clade_name)) %>%
  dplyr::filter(!grepl("unclassified", clade_name, ignore.case = TRUE))

phylum_names <- gsub(".*p__", "", phylum_df$clade_name)
phylum_abund <- as.data.frame(phylum_df[, colnames(phylum_df) %in% shared_samples])
rownames(phylum_abund) <- phylum_names

# Pivot to long format
phylum_long <- phylum_abund %>%
  rownames_to_column("Phylum") %>%
  pivot_longer(-Phylum, names_to = "sample_id", values_to = "Abundance") %>%
  dplyr::left_join(metadata[, c("sample_id", "source")], by = "sample_id")

# Order samples by group
sample_order <- metadata %>%
  dplyr::arrange(source, sample_id) %>%
  dplyr::pull(sample_id)

phylum_long$sample_id <- factor(phylum_long$sample_id, levels = sample_order)

# Colour palette
n_phyla <- length(unique(phylum_long$Phylum))
phylum_cols <- colorRampPalette(brewer.pal(min(n_phyla, 12), "Set3"))(n_phyla)

ggplot(phylum_long, aes(x = sample_id, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", width = 0.85) +
  scale_fill_manual(values = phylum_cols) +
  labs(x = NULL, y = "Relative Abundance (%)") +
  facet_grid(~ source, scales = "free_x", space = "free_x") +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  ) +
  guides(fill = guide_legend(ncol = 3))
```

\newpage

# Differential Abundance Analysis

MetaPhlAn4 outputs relative abundances (percentages). Both DESeq2 and edgeR expect integer count data, so we generate pseudo-counts by scaling TSS-normalised proportions by $10^6$ and rounding. Each method then applies its own internal normalisation:

- **DESeq2:** median-of-ratios size factor estimation, which accounts for differences in library composition.
- **edgeR:** TMM (trimmed mean of M-values) normalisation, which adjusts for compositional biases between samples.

The >10% prevalence filter applied earlier (Section 1.7) is the only abundance-based filter; no additional filtering is applied here.

```{r pseudo-counts}
# Convert TSS proportions to pseudo-counts (scale to 1M reads per sample)
scaling_factor <- 1e6
count_mat_filt <- round(tss_mat * scaling_factor)
storage.mode(count_mat_filt) <- "integer"

cat("Count matrix for DA:", nrow(count_mat_filt), "taxa x", ncol(count_mat_filt), "samples\n")
```

## Helper: Top-N Differential Abundance Bar Plot

The function below is reused for both DESeq2 and edgeR. It takes the top `n_each` positive and top `n_each` negative significant taxa (by signed log fold change), then draws a horizontal bar plot of log fold change with bars filled by $-\log_{10}$(adjusted p-value).

```{r da-barplot-helper}
plot_da_bars <- function(df,
                         lfc_col,
                         padj_col,
                         label_col   = "Species",
                         padj_cutoff = 0.05,
                         n_each      = 10,
                         title       = NULL) {

  # Keep only rows with non-missing stats and significant adjusted p
  sig_df <- df %>%
    dplyr::filter(!is.na(.data[[padj_col]]),
                  !is.na(.data[[lfc_col]]),
                  .data[[padj_col]] < padj_cutoff)

  if (nrow(sig_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0,
                 label = "No taxa significant at the chosen threshold") +
        theme_void()
    )
  }

  top_pos <- sig_df %>%
    dplyr::filter(.data[[lfc_col]] > 0) %>%
    dplyr::arrange(dplyr::desc(.data[[lfc_col]])) %>%
    dplyr::slice_head(n = n_each)

  top_neg <- sig_df %>%
    dplyr::filter(.data[[lfc_col]] < 0) %>%
    dplyr::arrange(.data[[lfc_col]]) %>%
    dplyr::slice_head(n = n_each)

  plot_df <- dplyr::bind_rows(top_pos, top_neg)
  plot_df$neg_log10_padj <- -log10(plot_df[[padj_col]])
  # Order taxa on the y-axis by signed log fold change
  plot_df[[label_col]] <- factor(
    plot_df[[label_col]],
    levels = plot_df[[label_col]][order(plot_df[[lfc_col]])]
  )

  ggplot(plot_df,
         aes(x = .data[[lfc_col]],
             y = .data[[label_col]],
             fill = neg_log10_padj)) +
    geom_col(width = 0.75) +
    geom_vline(xintercept = 0, colour = "black", linewidth = 0.4) +
    scale_fill_viridis_c(
      option = "plasma",
      name   = expression(-Log[10]~"(Adj. P-value)")
    ) +
    labs(
      title = title,
      x = expression(Log[2]~"Fold Change"),
      y = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(
      plot.title   = element_text(hjust = 0.5, face = "bold"),
      axis.text.y  = element_text(size = 9, face = "italic"),
      panel.grid.major.y = element_blank(),
      legend.position = "right",
      legend.key.height = unit(0.8, "cm")
    )
}
```

## DESeq2

```{r deseq2}
dds <- DESeqDataSetFromMatrix(
  countData = count_mat_filt,
  colData   = metadata,
  design    = ~ source
)

# Run DESeq2 with default settings
dds <- DESeq(dds)

# Extract results: HRSSL vs NN (positive log2FC = higher in HRSSL)
res_deseq2 <- results(dds, contrast = c("source", "HRSSL", "NN"), alpha = 0.05)
res_deseq2_df <- as.data.frame(res_deseq2) %>%
  rownames_to_column("Species") %>%
  dplyr::arrange(padj) %>%
  dplyr::mutate(
    Direction = dplyr::case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "HRSSL-enriched",
      padj < 0.05 & log2FoldChange < 0 ~ "NN-enriched",
      TRUE ~ "NS"
    )
  )

cat("DESeq2 significant taxa (padj < 0.05):", sum(res_deseq2_df$padj < 0.05, na.rm = TRUE), "\n")
```

### DESeq2 Results Table (Top 50 by Adjusted P-value)

The table below shows the top 50 taxa ranked by adjusted p-value. The full result set (all tested taxa) is retained in `res_deseq2_df` for the downstream consensus and log fold change comparison sections.

```{r deseq2-table}
kable(
  res_deseq2_df %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = 50) %>%
    dplyr::select(Species, baseMean, log2FoldChange, lfcSE, pvalue, padj, Direction) %>%
    dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4))),
  booktabs = TRUE, longtable = TRUE,
  caption = "Top 50 DESeq2 differential abundance results, ranked by adjusted p-value (HRSSL vs NN)."
) %>%
  kable_styling(
    latex_options = c("repeat_header", "scale_down", "HOLD_position"),
    font_size = 8
  )
```

### DESeq2 Top Differentially Abundant Taxa

```{r deseq2-da-barplot, fig.height=6, fig.width=8, fig.cap="DESeq2: top 10 HRSSL-enriched (log2FC > 0) and top 10 NN-enriched (log2FC < 0) taxa at padj < 0.05, ranked by signed log2 fold change. Bars are filled by -log10(adjusted p-value)."}
plot_da_bars(
  df          = res_deseq2_df,
  lfc_col     = "log2FoldChange",
  padj_col    = "padj",
  label_col   = "Species",
  padj_cutoff = 0.05,
  n_each      = 10,
  title       = "DESeq2: NN vs HRSSL"
)
```

### DESeq2 Volcano Plot

```{r deseq2-volcano, fig.height=5, fig.cap="DESeq2 volcano plot. Red = HRSSL-enriched, blue = NN-enriched (padj < 0.05)."}
ggplot(res_deseq2_df, aes(x = log2FoldChange, y = -log10(pvalue), colour = Direction)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_manual(values = c("HRSSL-enriched" = "#E41A1C",
                                  "NN-enriched"    = "#377EB8",
                                  "NS"             = "grey60")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  ggrepel::geom_text_repel(
    data = dplyr::filter(res_deseq2_df, padj < 0.05),
    aes(label = Species), size = 3, max.overlaps = 15, show.legend = FALSE
  ) +
  labs(x = "log2 Fold Change (HRSSL / NN)", y = "-log10(p-value)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

\newpage

## edgeR

```{r edger}
# Create DGEList (prevalence filter already applied at Section 1.7)
dge <- DGEList(counts = count_mat_filt, group = metadata$source)

cat("Taxa input to edgeR:", nrow(dge), "\n")

# Normalisation (TMM)
dge <- calcNormFactors(dge, method = "TMM")

# Design matrix
design_mat <- model.matrix(~ source, data = metadata)

# Estimate dispersions
dge <- estimateDisp(dge, design_mat)

# Quasi-likelihood F-test (robust)
fit <- glmQLFit(dge, design_mat)
qlf <- glmQLFTest(fit, coef = "sourceHRSSL")

# Extract results
res_edger <- topTags(qlf, n = Inf, sort.by = "PValue")$table
res_edger_df <- as.data.frame(res_edger) %>%
  rownames_to_column("Species") %>%
  dplyr::mutate(
    Direction = dplyr::case_when(
      FDR < 0.05 & logFC > 0 ~ "HRSSL-enriched",
      FDR < 0.05 & logFC < 0 ~ "NN-enriched",
      TRUE ~ "NS"
    )
  )

cat("edgeR significant taxa (FDR < 0.05):", sum(res_edger_df$FDR < 0.05, na.rm = TRUE), "\n")
```

### edgeR Results Table (Top 50 by FDR)

The table below shows the top 50 taxa ranked by FDR. The full result set is retained in `res_edger_df` for downstream sections.

```{r edger-table}
kable(
  res_edger_df %>%
    dplyr::arrange(FDR) %>%
    dplyr::slice_head(n = 50) %>%
    dplyr::select(Species, logFC, logCPM, F, PValue, FDR, Direction) %>%
    dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4))),
  booktabs = TRUE, longtable = TRUE,
  caption = "Top 50 edgeR quasi-likelihood F-test results, ranked by FDR (HRSSL vs NN)."
) %>%
  kable_styling(
    latex_options = c("repeat_header", "scale_down", "HOLD_position"),
    font_size = 8
  )
```

### edgeR Top Differentially Abundant Taxa

```{r edger-da-barplot, fig.height=6, fig.width=8, fig.cap="edgeR: top 10 HRSSL-enriched (logFC > 0) and top 10 NN-enriched (logFC < 0) taxa at FDR < 0.05, ranked by signed log fold change. Bars are filled by -log10(FDR)."}
plot_da_bars(
  df          = res_edger_df,
  lfc_col     = "logFC",
  padj_col    = "FDR",
  label_col   = "Species",
  padj_cutoff = 0.05,
  n_each      = 10,
  title       = "edgeR: NN vs HRSSL"
)
```

### edgeR Volcano Plot

```{r edger-volcano, fig.height=5, fig.cap="edgeR volcano plot. Red = HRSSL-enriched, blue = NN-enriched (FDR < 0.05)."}
ggplot(res_edger_df, aes(x = logFC, y = -log10(PValue), colour = Direction)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_manual(values = c("HRSSL-enriched" = "#E41A1C",
                                  "NN-enriched"    = "#377EB8",
                                  "NS"             = "grey60")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  ggrepel::geom_text_repel(
    data = dplyr::filter(res_edger_df, FDR < 0.05),
    aes(label = Species), size = 3, max.overlaps = 15, show.legend = FALSE
  ) +
  labs(x = "logFC (HRSSL / NN)", y = "-log10(p-value)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

\newpage

# Consensus Between DESeq2 and edgeR

```{r consensus}
sig_deseq2 <- res_deseq2_df %>%
  dplyr::filter(padj < 0.05) %>%
  dplyr::select(Species, log2FoldChange_DESeq2 = log2FoldChange, padj_DESeq2 = padj)

sig_edger <- res_edger_df %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::select(Species, logFC_edgeR = logFC, FDR_edgeR = FDR)

consensus <- dplyr::inner_join(sig_deseq2, sig_edger, by = "Species")

if (nrow(consensus) > 0) {
  cat("Consensus significant taxa (both methods, adjusted p < 0.05):", nrow(consensus), "\n\n")

  consensus_display <- consensus %>%
    dplyr::arrange(padj_DESeq2) %>%
    dplyr::transmute(
      Species,
      `log2FC (DESeq2)` = sprintf("%.2f", log2FoldChange_DESeq2),
      `padj (DESeq2)`   = format(padj_DESeq2, scientific = TRUE, digits = 3),
      `logFC (edgeR)`   = sprintf("%.2f", logFC_edgeR),
      `FDR (edgeR)`     = format(FDR_edgeR,  scientific = TRUE, digits = 3)
    )

  kable(
    consensus_display,
    booktabs  = TRUE,
    longtable = TRUE,
    align     = c("l", "r", "r", "r", "r"),
    caption   = "Taxa significant in both DESeq2 and edgeR (adjusted p < 0.05)."
  ) %>%
    kable_styling(
      latex_options = c("repeat_header", "HOLD_position"),
      font_size     = 8
    ) %>%
    column_spec(1, width = "5.5cm", italic = TRUE) %>%
    column_spec(2:5, width = "2.2cm")
} else {
  cat("No taxa reached significance (adjusted p < 0.05) in both methods.\n")
}
```

## Log Fold Change Comparison

```{r lfc-comparison, fig.height=5, fig.cap="Comparison of log fold changes between DESeq2 and edgeR for all tested taxa."}
# Merge all results (not just significant)
lfc_compare <- dplyr::inner_join(
  res_deseq2_df %>% dplyr::select(Species, LFC_DESeq2 = log2FoldChange),
  res_edger_df  %>% dplyr::select(Species, LFC_edgeR  = logFC),
  by = "Species"
) |>
  dplyr::mutate(
    Concordance = dplyr::case_when(
      LFC_DESeq2 > 0 & LFC_edgeR > 0 ~ "Concordant increase",
      LFC_DESeq2 < 0 & LFC_edgeR < 0 ~ "Concordant decrease",
      TRUE ~ "Discordant"
    )
  )

if (nrow(lfc_compare) > 1) {
  r_val <- cor(lfc_compare$LFC_DESeq2, lfc_compare$LFC_edgeR,
               use = "complete.obs", method = "spearman")

  ggplot(lfc_compare, aes(x = LFC_DESeq2, y = LFC_edgeR, colour = Concordance)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    ggrepel::geom_text_repel(aes(label = Species), size = 2.5, max.overlaps = 10,
                              show.legend = FALSE) +
    annotate("text", x = min(lfc_compare$LFC_DESeq2, na.rm = TRUE),
             y = max(lfc_compare$LFC_edgeR, na.rm = TRUE),
             label = paste0("Spearman r = ", round(r_val, 3)),
             hjust = 0, size = 4, colour = "black") +
    scale_colour_manual(values = c(
      "Concordant increase" = "#D55E00",
      "Concordant decrease" = "#0072B2",
      "Discordant"          = "grey60"
    )) +
    labs(x = "DESeq2 log2FoldChange", y = "edgeR logFC", colour = "Direction") +
    theme_bw(base_size = 12)
} else {
  cat("Insufficient overlapping taxa to compare fold changes.\n")
}
```

\newpage

# Session Info

```{r session-info}
sessionInfo()
```
