---
title: "NN - No Neoplasia: FIT vs Fresh Collection — Microbiome Comparability Analysis"
subtitle: "Demonstrating comparable metagenomic profiles across faecal collection methods"
author: "Brad Hart"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
    keep_tex: false
    includes:
      in_header: null
geometry: "margin=2.5cm"
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos  = "H",
  dpi     = 300
)
```

```{r load-packages, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(phyloseq)
  library(vegan)
  library(ggpubr)
  library(patchwork)
  library(ALDEx2)
  library(dplyr)
  library(knitr)
  library(kableExtra)
})

set.seed(42)
```

```{r theme-setup, include=FALSE}
theme_pub <- theme_bw(base_size = 12) +
  theme(
    panel.grid.minor   = element_blank(),
    strip.background   = element_rect(fill = "grey95", colour = NA),
    legend.position    = "bottom",
    plot.title         = element_text(face = "bold", size = 13),
    axis.title         = element_text(size = 11),
    legend.title       = element_text(size = 10),
    legend.text        = element_text(size = 9)
  )

group_cols <- c("direct" = "#4E79A7", "FIT" = "#E15759")
# Display labels only: the "direct" group is shown as "fresh" in figure
# legends and axis text; the underlying metadata values are unchanged.
group_labs <- c("direct" = "fresh", "FIT" = "FIT")
```

---

# Overview

This report compares faecal microbiome profiles from two collection methods --- **FIT** and **fresh collection** --- using whole-genome shotgun metagenomic data profiled with **MetaPhlAn4**. The goal is to demonstrate that FIT-based collection does not introduce systematic bias relative to fresh sampling, supporting its use as a suitable option for genomic biomarker identification methods.

The No Neoplasia (NN) fresh collection samples were obtained from the supplementary material of the following publication:

> Piccinno, G., Thompson, K.N., Manghi, P. *et al.* (2025). Pooled analysis of 3,741 stool metagenomes from 18 cohorts for cross-stage and strain-level reproducible microbial biomarkers of colorectal cancer. *Nature Medicine*, 31, 2416--2429. <https://doi.org/10.1038/s41591-025-03693-9>

**Important caveat:** The two groups come from **different individuals on different continents**. Any detected differences may therefore reflect geography, diet, or host genetics rather than collection method. Results are interpreted conservatively with this confound in mind.

---

# Data Import & Preprocessing

```{r read-abundance, include=FALSE}
abund_raw <- read_tsv(
  "merged_abundance_table_natureandNN.tsv",
  show_col_types = FALSE
)
```

```{r filter-deepest-rank, include=FALSE}
deepest_rank <- if (any(str_detect(abund_raw$clade_name, "s__"))) {
  "s__"
} else {
  ranks <- c("t__", "s__", "g__", "f__", "o__", "c__", "p__", "k__")
  detected <- ranks[sapply(ranks, function(r) any(str_detect(abund_raw$clade_name, fixed(r))))]
  detected[1]
}

if (deepest_rank == "s__") {
  abund_filt <- abund_raw %>%
    filter(str_detect(clade_name, "s__"), !str_detect(clade_name, "t__"))
} else {
  all_clades <- abund_raw$clade_name
  is_leaf <- sapply(all_clades, function(x) {
    !any(str_detect(all_clades, fixed(paste0(x, "|"))))
  })
  abund_filt <- abund_raw[is_leaf, ]
}
```

```{r parse-taxonomy, include=FALSE}
tax_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

parse_metaphlan_taxonomy <- function(clade_names) {
  prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__")
  tax_df <- matrix(NA_character_, nrow = length(clade_names), ncol = 7,
                   dimnames = list(NULL, tax_ranks))
  for (i in seq_along(prefixes)) {
    tax_df[, i] <- str_extract(clade_names, paste0(prefixes[i], "[^|]+")) %>%
      str_remove(fixed(prefixes[i]))
  }
  as.data.frame(tax_df, stringsAsFactors = FALSE)
}

tax_table_df <- parse_metaphlan_taxonomy(abund_filt$clade_name)

# Build meaningful rownames: prefer Species > Genus > Family > placeholder
taxon_labels <- tax_table_df$Species
taxon_labels[is.na(taxon_labels)] <- tax_table_df$Genus[is.na(taxon_labels)]
taxon_labels[is.na(taxon_labels)] <- tax_table_df$Family[is.na(taxon_labels)]
taxon_labels[is.na(taxon_labels)] <- paste0("unclassified_", seq_along(which(is.na(taxon_labels))))
taxon_labels <- str_replace_all(taxon_labels, "_", " ")
rownames(tax_table_df) <- make.unique(taxon_labels)
```

```{r build-matrices, include=FALSE}
select <- dplyr::select
otu_mat <- abund_filt %>%
  select(-clade_name) %>%
  as.data.frame()
rownames(otu_mat) <- rownames(tax_table_df)

meta <- read_tsv("metadata_fit_vs_direct.tsv", show_col_types = FALSE) %>%
  column_to_rownames("sample_id")
meta$source <- factor(meta$source, levels = c("direct", "FIT"))

shared_samples <- intersect(colnames(otu_mat), rownames(meta))
otu_mat <- otu_mat[, shared_samples]
meta    <- meta[shared_samples, , drop = FALSE]
```

```{r build-phyloseq, include=FALSE}
ps <- phyloseq(
  otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_table_df)),
  sample_data(meta)
)
```

```{r prevalence-filter, include=FALSE}
prevalence_threshold <- 0.10
min_samples <- ceiling(prevalence_threshold * nsamples(ps))
taxa_prev   <- apply(otu_table(ps), 1, function(x) sum(x > 0))
keep_taxa   <- names(taxa_prev[taxa_prev >= min_samples])
ps_filt     <- prune_taxa(keep_taxa, ps)
```

```{r normalise, include=FALSE}
ps_rel     <- transform_sample_counts(ps_filt, function(x) x / sum(x) * 100)
pseudo     <- 0.5
otu_pseudo <- as(otu_table(ps_filt), "matrix") + pseudo
clr_mat    <- t(apply(otu_pseudo, 2, function(x) log(x) - mean(log(x))))
```

The MetaPhlAn4 merged abundance table was filtered to the deepest taxonomic rank detected (`r deepest_rank`). A minimum prevalence filter of 10% of samples was applied, retaining `r ntaxa(ps_filt)` of `r ntaxa(ps)` taxa across `r nsamples(ps_filt)` samples (`r sum(meta$source == "direct")` fresh, `r sum(meta$source == "FIT")` FIT). Sample group sizes:

```{r sample-summary-table}
sample_counts <- as.data.frame(table(meta$source)) %>%
  rename(Collection_Method = Var1, N_Samples = Freq) %>%
  # Display label only; the underlying metadata value is unchanged.
  mutate(Collection_Method = unname(group_labs[as.character(Collection_Method)]))

kable(
  sample_counts,
  col.names = c("Collection Method", "N Samples"),
  align     = c("l", "c"),
  booktabs  = TRUE,
  caption   = "Sample counts by collection method."
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

---

# Alpha Diversity

Alpha diversity measures within-sample richness and evenness. Shannon and Simpson indices were compared between collection methods using the Wilcoxon rank-sum test. Cliff's d was calculated as a non-parametric effect size (|d| < 0.147 = negligible, < 0.33 = small).

```{r alpha-compute, include=FALSE}
ps_counts <- transform_sample_counts(ps_filt, function(x) round(x / sum(x) * 10000))

alpha_df <- estimate_richness(ps_counts, measures = c("Shannon", "Simpson")) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta %>% rownames_to_column("sample_id"), by = "sample_id")

shannon_test <- wilcox.test(Shannon ~ source, data = alpha_df, exact = FALSE)
simpson_test <- wilcox.test(Simpson ~ source, data = alpha_df, exact = FALSE)

cliffs_delta <- function(x, y) {
  comparisons <- outer(x, y, FUN = function(a, b) sign(a - b))
  sum(comparisons) / (length(x) * length(y))
}

cd_shannon <- cliffs_delta(
  alpha_df$Shannon[alpha_df$source == "direct"],
  alpha_df$Shannon[alpha_df$source == "FIT"]
)
cd_simpson <- cliffs_delta(
  alpha_df$Simpson[alpha_df$source == "direct"],
  alpha_df$Simpson[alpha_df$source == "FIT"]
)
```

```{r alpha-stats-table}
alpha_stats <- tibble(
  Index       = c("Shannon", "Simpson"),
  W_statistic = c(round(shannon_test$statistic, 1), round(simpson_test$statistic, 1)),
  p_value     = c(round(shannon_test$p.value, 4), round(simpson_test$p.value, 4)),
  Cliffs_d    = c(round(cd_shannon, 3), round(cd_simpson, 3)),
  Result      = c(
    ifelse(shannon_test$p.value > 0.05, "Not significant", "Significant"),
    ifelse(simpson_test$p.value > 0.05, "Not significant", "Significant")
  )
)

kable(
  alpha_stats,
  col.names = c("Index", "W statistic", "p-value", "Cliff's d", "Result"),
  align     = c("l", "c", "c", "c", "l"),
  booktabs  = TRUE,
  caption   = "Alpha diversity: Wilcoxon rank-sum test results."
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r alpha-plot, fig.width=7, fig.height=4.5, out.width="\\linewidth", fig.cap="Alpha diversity (Shannon and Simpson) by collection method. Boxes show median and IQR; points are individual samples."}
p_shannon <- ggplot(alpha_df, aes(x = source, y = Shannon, fill = source)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 2.2, alpha = 0.8, shape = 21, colour = "black",
              aes(fill = source)) +
  scale_fill_manual(values = group_cols, labels = group_labs) +
  scale_x_discrete(labels = group_labs) +
  labs(
    title    = "Shannon Diversity",
    subtitle = sprintf("Wilcoxon p = %.3f | Cliff's d = %.2f",
                       shannon_test$p.value, cd_shannon),
    x = "Collection method", y = "Shannon index"
  ) +
  theme_pub + theme(legend.position = "none")

p_simpson <- ggplot(alpha_df, aes(x = source, y = Simpson, fill = source)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 2.2, alpha = 0.8, shape = 21, colour = "black",
              aes(fill = source)) +
  scale_fill_manual(values = group_cols, labels = group_labs) +
  scale_x_discrete(labels = group_labs) +
  labs(
    title    = "Simpson Diversity",
    subtitle = sprintf("Wilcoxon p = %.3f | Cliff's d = %.2f",
                       simpson_test$p.value, cd_simpson),
    x = "Collection method", y = "Simpson index"
  ) +
  theme_pub + theme(legend.position = "none")

p_alpha <- p_shannon + p_simpson +
  plot_annotation(
    title    = "Alpha Diversity: FIT vs Fresh Collection",
    subtitle = "No significant difference in within-sample diversity",
    theme    = theme(plot.title    = element_text(face = "bold", size = 14),
                     plot.subtitle = element_text(size = 11, colour = "grey40"))
  )

p_alpha
```

**Interpretation:** Non-significant Wilcoxon tests and Cliff's d values near zero indicate that within-sample diversity is comparable between FIT and fresh collection methods. FIT collection does not systematically reduce or inflate microbial richness or evenness relative to fresh sampling.

---

# Beta Diversity

Beta diversity quantifies between-sample compositional differences via PCoA ordination using two complementary distance metrics.

```{r beta-distances, include=FALSE}
bc_dist  <- vegdist(t(as(otu_table(ps_rel), "matrix")), method = "bray")
ait_dist <- dist(clr_mat, method = "euclidean")

pcoa_bc    <- cmdscale(bc_dist, eig = TRUE, k = 2)
var_bc     <- round(pcoa_bc$eig[1:2] / sum(pcoa_bc$eig[pcoa_bc$eig > 0]) * 100, 1)
pcoa_bc_df <- data.frame(
  PCo1   = pcoa_bc$points[, 1],
  PCo2   = pcoa_bc$points[, 2],
  source = meta[rownames(pcoa_bc$points), "source"]
)

pcoa_ait    <- cmdscale(ait_dist, eig = TRUE, k = 2)
var_ait     <- round(pcoa_ait$eig[1:2] / sum(pcoa_ait$eig[pcoa_ait$eig > 0]) * 100, 1)
pcoa_ait_df <- data.frame(
  PCo1   = pcoa_ait$points[, 1],
  PCo2   = pcoa_ait$points[, 2],
  source = meta[rownames(pcoa_ait$points), "source"]
)
```

```{r beta-plot, fig.width=7, fig.height=5, out.width="\\linewidth", fig.cap="PCoA ordination using Bray--Curtis (left) and Aitchison (right) distances. Dashed ellipses represent 95\\% confidence regions per group."}
p_pcoa_bc <- ggplot(pcoa_bc_df, aes(x = PCo1, y = PCo2, colour = source)) +
  geom_point(size = 3, alpha = 0.85) +
  stat_ellipse(level = 0.95, linetype = "dashed", linewidth = 0.7) +
  scale_colour_manual(values = group_cols, labels = group_labs) +
  labs(
    title  = "PCoA \u2014 Bray\u2013Curtis",
    x      = paste0("PCo1 (", var_bc[1], "%)"),
    y      = paste0("PCo2 (", var_bc[2], "%)"),
    colour = "Collection method"
  ) +
  theme_pub

p_pcoa_ait <- ggplot(pcoa_ait_df, aes(x = PCo1, y = PCo2, colour = source)) +
  geom_point(size = 3, alpha = 0.85) +
  stat_ellipse(level = 0.95, linetype = "dashed", linewidth = 0.7) +
  scale_colour_manual(values = group_cols, labels = group_labs) +
  labs(
    title  = "PCoA \u2014 Aitchison",
    x      = paste0("PCo1 (", var_ait[1], "%)"),
    y      = paste0("PCo2 (", var_ait[2], "%)"),
    colour = "Collection method"
  ) +
  theme_pub

p_beta <- p_pcoa_bc + p_pcoa_ait +
  plot_annotation(
    title    = "Beta Diversity: FIT vs Fresh Collection",
    subtitle = "Overlapping confidence ellipses indicate shared community structure",
    theme    = theme(plot.title    = element_text(face = "bold", size = 14),
                     plot.subtitle = element_text(size = 11, colour = "grey40"))
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_beta
```

**Interpretation:** Overlapping 95% confidence ellipses indicate that FIT and fresh samples occupy largely shared ordination space. Inter-individual and geographic variation is expected to dominate over any collection-method effect. The absence of distinct, well-separated clusters is consistent with FIT collection preserving the overall structure of faecal microbiome composition.

---

# PERMANOVA

PERMANOVA (`adonis2`) partitions total community variation and tests whether collection method explains a significant fraction, reported as R2 (effect size) and permutation p-value. **Important caveat:** because samples originate from **different individuals on different continents**, any significant R2 attributed to collection method must be interpreted with caution --- the geographic confound means that observed differences may reflect population-level biology rather than collection-method bias.

```{r permanova, include=FALSE}
set.seed(42)
perm_bc  <- adonis2(bc_dist ~ source, data = meta, permutations = 999)

set.seed(42)
perm_ait <- adonis2(ait_dist ~ source, data = meta, permutations = 999)
```

```{r permanova-table}
perm_table <- tibble(
  Distance    = c("Bray-Curtis", "Aitchison"),
  F_statistic = c(round(perm_bc$F[1], 2), round(perm_ait$F[1], 2)),
  R2          = c(round(perm_bc$R2[1], 4), round(perm_ait$R2[1], 4)),
  Pct_var     = paste0(round(c(perm_bc$R2[1], perm_ait$R2[1]) * 100, 1), "%"),
  p_value     = c(perm_bc$`Pr(>F)`[1], perm_ait$`Pr(>F)`[1]),
  Result      = ifelse(
    c(perm_bc$`Pr(>F)`[1], perm_ait$`Pr(>F)`[1]) > 0.05,
    "Not significant", "Significant"
  )
)

kable(
  perm_table,
  col.names = c("Distance metric", "F statistic", "R2", "Variance explained", "p-value", "Result"),
  align     = c("l", "c", "c", "c", "c", "l"),
  booktabs  = TRUE,
  caption   = "PERMANOVA (adonis2, 999 permutations) results by distance metric."
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  column_spec(1, width = "2.5cm") %>%
  column_spec(2, width = "1.5cm") %>%
  column_spec(3, width = "1cm") %>%
  column_spec(4, width = "2cm") %>%
  column_spec(5, width = "1.2cm") %>%
  column_spec(6, width = "2cm")
```

**Interpretation:** R2 indicates the fraction of total variation attributable to collection method. A small R2 (< 5--10%) provides evidence that collection method explains negligible compositional variation. Even if p < 0.05, a small R2 means the effect is minor --- and given the geographic confound, any significance is more likely to reflect continent-of-origin differences than a genuine collection-method bias.

---

# PERMDISP

PERMDISP tests whether within-group dispersion (spread around each group's centroid) differs between FIT and fresh, validating a key PERMANOVA assumption.

```{r permdisp, include=FALSE}
bd_bc  <- betadisper(bc_dist, meta$source)
pd_bc  <- permutest(bd_bc, permutations = 999)

bd_ait <- betadisper(ait_dist, meta$source)
pd_ait <- permutest(bd_ait, permutations = 999)
```

```{r permdisp-table}
permdisp_table <- tibble(
  Distance    = c("Bray-Curtis", "Aitchison"),
  F_statistic = c(round(pd_bc$tab$F[1], 2), round(pd_ait$tab$F[1], 2)),
  p_value     = c(pd_bc$tab$`Pr(>F)`[1], pd_ait$tab$`Pr(>F)`[1]),
  Result      = ifelse(
    c(pd_bc$tab$`Pr(>F)`[1], pd_ait$tab$`Pr(>F)`[1]) > 0.05,
    "Not significant", "Significant"
  )
)

kable(
  permdisp_table,
  col.names = c("Distance metric", "F statistic", "p-value", "Result"),
  align     = c("l", "c", "c", "l"),
  booktabs  = TRUE,
  caption   = "PERMDISP (betadisper, 999 permutations): homogeneity of group dispersions."
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r permdisp-plot, fig.width=7, fig.height=4.5, out.width="\\linewidth", fig.cap="PERMDISP: distance of each sample to its group centroid, split by distance metric. Similar distributions indicate homogeneous within-group dispersion."}
disp_df <- bind_rows(
  tibble(distance = bd_bc$distances,  source = meta$source, metric = "Bray\u2013Curtis"),
  tibble(distance = bd_ait$distances, source = meta$source, metric = "Aitchison")
)

p_permdisp <- ggplot(disp_df, aes(x = source, y = distance, fill = source)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.7, shape = 21, colour = "black",
              aes(fill = source)) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = group_cols, labels = group_labs) +
  scale_x_discrete(labels = group_labs) +
  labs(
    title    = "PERMDISP: Homogeneity of Group Dispersions",
    subtitle = "Similar spread around centroids confirms comparable within-group variability",
    x = "Collection method", y = "Distance to group centroid"
  ) +
  theme_pub + theme(legend.position = "none")

p_permdisp
```

**Interpretation:** Non-significant PERMDISP results confirm that within-group dispersion is homogeneous --- FIT samples are not more variable around their centroid than fresh samples, and vice versa. This validates the PERMANOVA assumption and provides additional evidence that FIT collection does not introduce excess microbiome variability.

---

# Differential Abundance (ALDEx2)

ALDEx2 uses a Bayesian framework with Monte Carlo sampling from a Dirichlet distribution, making it robust to compositionality and conservative regarding false positives. Benjamini--Hochberg correction was applied. Note that because the two groups originate from different cohorts, any significant taxa should be interpreted cautiously --- the signal may partly reflect population-level differences rather than a genuine collection-method effect.

```{r aldex2-run, include=FALSE}
otu_aldex <- as(otu_table(ps_filt), "matrix")
if (max(otu_aldex) <= 100) {
  otu_aldex <- round(otu_aldex / colSums(otu_aldex) * 10000)
}
mode(otu_aldex) <- "integer"

conditions <- as.character(meta[colnames(otu_aldex), "source"])

set.seed(42)
aldex_res <- aldex(
  reads      = otu_aldex,
  conditions = conditions,
  mc.samples = 128,
  test       = "t",
  effect     = TRUE,
  denom      = "all",
  verbose    = FALSE
)

sig_aldex <- aldex_res %>%
  rownames_to_column("taxon") %>%
  filter(wi.eBH < 0.05 | we.eBH < 0.05)
```

```{r aldex2-plot, fig.width=6.5, fig.height=5, fig.cap="ALDEx2 effect plot. x-axis: within-group dispersion; y-axis: between-group difference. Points near y = 0 indicate no meaningful difference between collection methods."}
aldex_plot_df <- aldex_res %>%
  rownames_to_column("taxon") %>%
  mutate(significant = factor(
    ifelse(wi.eBH < 0.05 | we.eBH < 0.05,
           "Significant (BH p < 0.05)", "Not significant"),
    levels = c("Not significant", "Significant (BH p < 0.05)")
  ))

p_aldex <- ggplot(aldex_plot_df, aes(x = diff.win, y = diff.btw, colour = significant)) +
  geom_point(size = 2.5, alpha = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Not significant"           = "grey65",
                                 "Significant (BH p < 0.05)" = "#E15759")) +
  labs(
    title    = "ALDEx2: Differential Abundance",
    subtitle = "Points near y = 0 indicate no between-group difference",
    x        = "Dispersion (within-group difference)",
    y        = "Effect (between-group difference)",
    colour   = NULL
  ) +
  theme_pub

p_aldex
```

```{r aldex2-sig-table, results='asis'}
if (nrow(sig_aldex) == 0) {
  cat("\\textbf{No taxa were significantly differentially abundant} between FIT and fresh collection after Benjamini--Hochberg correction (", nrow(aldex_res), " taxa tested).", sep = "")
} else {
  tax_lookup <- tax_table_df %>%
    rownames_to_column("taxon") %>%
    select(taxon, Genus, Species)

  sig_display <- sig_aldex %>%
    left_join(tax_lookup, by = "taxon") %>%
    mutate(
      Species_label = case_when(
        !is.na(Species) ~ cell_spec(str_replace_all(Species, "_", " "),
                                    format = "latex", italic = TRUE),
        !is.na(Genus)   ~ cell_spec(paste0(str_replace_all(Genus, "_", " "), " sp."),
                                    format = "latex", italic = TRUE),
        TRUE            ~ taxon
      ),
      Genus_label = ifelse(!is.na(Genus),
                           cell_spec(str_replace_all(Genus, "_", " "),
                                     format = "latex", italic = TRUE),
                           "Unknown")
    ) %>%
    select(Species_label, Genus_label, diff.btw, diff.win, effect, wi.eBH, we.eBH) %>%
    arrange(wi.eBH) %>%
    mutate(across(c(diff.btw, diff.win, effect, wi.eBH, we.eBH), ~round(.x, 4)))

  kable(
    sig_display,
    col.names  = c("Species", "Genus", "Diff between", "Diff within",
                   "Effect size", "Wilcoxon BH q", "Welch BH q"),
    booktabs   = TRUE,
    escape     = FALSE,
    caption    = paste0("Significantly differentially abundant taxa (BH-adjusted p < 0.05). ",
                        nrow(sig_aldex), " of ", nrow(aldex_res), " taxa tested.")
  ) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9) %>%
    column_spec(1, width = "3.5cm") %>%
    column_spec(2, width = "2.5cm") %>%
    column_spec(3, width = "1.5cm") %>%
    column_spec(4, width = "1.5cm") %>%
    column_spec(5, width = "1.5cm") %>%
    column_spec(6, width = "1.8cm") %>%
    column_spec(7, width = "1.5cm")
}
```

**Interpretation:** ALDEx2's conservative Bayesian framework with BH correction guards against false positives. `r if (nrow(sig_aldex) == 0) "No taxa are differentially abundant, strongly supporting the conclusion that FIT collection does not bias individual taxon abundances relative to fresh collection." else paste0(nrow(sig_aldex), " taxa reached significance after BH correction. Given the geographic confound, these findings should be interpreted cautiously --- the signal may partly reflect population-level differences between cohorts rather than a genuine collection-method effect.")` Points near the y = 0 line confirm that the large majority of taxa show no systematic between-group difference.

---

# Taxonomic Abundance Profiles

The most abundant taxa are summarised below at species, genus, and family level. For each rank the ten taxa with the highest overall mean relative abundance are listed, and the mean relative abundance is shown separately for fresh and FIT collection so the two methods can be compared directly. A collection method that preserves composition should return a similar ranking and similar mean abundances across the two columns.

```{r abundance-compute, include=FALSE}
top_abund_table <- function(ps, rank, n = 10) {
  ps_glom <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  melted  <- psmelt(ps_glom) %>%
    mutate(taxon = .data[[rank]],
           taxon = ifelse(is.na(taxon) | taxon == "", "Unclassified", taxon))

  overall <- melted %>%
    group_by(taxon) %>%
    summarise(Overall = mean(Abundance), .groups = "drop")

  bygroup <- melted %>%
    group_by(taxon, source) %>%
    summarise(mean_ab = mean(Abundance), .groups = "drop") %>%
    pivot_wider(names_from = source, values_from = mean_ab, values_fill = 0)

  overall %>%
    left_join(bygroup, by = "taxon") %>%
    arrange(desc(Overall)) %>%
    slice_head(n = n)
}

render_abund_kable <- function(tab, rank_label, caption) {
  tab %>%
    mutate(
      Taxon   = cell_spec(str_replace_all(taxon, "_", " "),
                          format = "latex", italic = TRUE),
      Direct  = round(direct, 3),
      FIT     = round(FIT, 3),
      Overall = round(Overall, 3)
    ) %>%
    select(Taxon, Direct, FIT, Overall) %>%
    kable(
      col.names = c(rank_label, "Fresh (mean %)", "FIT (mean %)", "Overall (mean %)"),
      align     = c("l", "c", "c", "c"),
      booktabs  = TRUE,
      escape    = FALSE,
      caption   = caption
    ) %>%
    kable_styling(latex_options = c("hold_position"), font_size = 9) %>%
    column_spec(1, width = "6cm")
}

species_tab <- top_abund_table(ps_rel, "Species")
genus_tab   <- top_abund_table(ps_rel, "Genus")
family_tab  <- top_abund_table(ps_rel, "Family")
```

## Species-Level Abundance

```{r abundance-species}
render_abund_kable(
  species_tab,
  "Species",
  "Top 10 species by overall mean relative abundance, with per-method means."
)
```

## Genus-Level Abundance

```{r abundance-genus}
render_abund_kable(
  genus_tab,
  "Genus",
  "Top 10 genera by overall mean relative abundance, with per-method means."
)
```

## Family-Level Abundance

```{r abundance-family}
render_abund_kable(
  family_tab,
  "Family",
  "Top 10 families by overall mean relative abundance, with per-method means."
)
```

## Taxa Detected by Collection Method

The table below reports the number of distinct taxa detected under each collection method (relative abundance above zero), together with the count shared by both methods and the combined total. Counts reflect the prevalence-filtered set used throughout this report.

```{r taxa-count-compute, include=FALSE}
taxa_by_method <- function(ps, rank) {
  ps_glom <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  m   <- as(otu_table(ps_glom), "matrix")
  if (!taxa_are_rows(ps_glom)) m <- t(m)
  src <- as.character(sample_data(ps_glom)$source)

  direct_present <- rowSums(m[, src == "direct", drop = FALSE] > 0) > 0
  fit_present    <- rowSums(m[, src == "FIT",    drop = FALSE] > 0) > 0

  tibble(
    Rank   = rank,
    Direct = sum(direct_present),
    FIT    = sum(fit_present),
    Shared = sum(direct_present & fit_present),
    Total  = sum(direct_present | fit_present)
  )
}

taxa_counts <- bind_rows(
  taxa_by_method(ps_rel, "Species"),
  taxa_by_method(ps_rel, "Genus"),
  taxa_by_method(ps_rel, "Family")
)
```

```{r taxa-count-table}
kable(
  taxa_counts,
  col.names = c("Rank", "Fresh", "FIT", "Shared", "Total (either)"),
  align     = c("l", "c", "c", "c", "c"),
  booktabs  = TRUE,
  caption   = "Number of distinct taxa detected by each collection method, with the count shared by both and the combined total."
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9)
```

**Interpretation:** A close correspondence between the fresh and FIT columns, in both the identity of the dominant taxa and their mean relative abundances, indicates that FIT collection preserves the abundance structure of the faecal community. Where a taxon ranks highly under one method but not the other, the cross-cohort design should be considered first, since geographic and dietary differences between the two populations can shift the abundance of individual taxa independently of collection method.

---

# Consolidated Results Summary

The table below consolidates all statistical results from this analysis.

```{r summary-table}
summary_table <- tribble(
  ~Analysis,                 ~Method,       ~Statistic,
  ~p_value,                  ~Effect_size,  ~Conclusion,

  "Alpha: Shannon",
    "Wilcoxon rank-sum",
    sprintf("W = %.1f", shannon_test$statistic),
    sprintf("%.4f", shannon_test$p.value),
    sprintf("Cliff's d = %.3f", cd_shannon),
    ifelse(shannon_test$p.value > 0.05, "Not significant", "Significant"),

  "Alpha: Simpson",
    "Wilcoxon rank-sum",
    sprintf("W = %.1f", simpson_test$statistic),
    sprintf("%.4f", simpson_test$p.value),
    sprintf("Cliff's d = %.3f", cd_simpson),
    ifelse(simpson_test$p.value > 0.05, "Not significant", "Significant"),

  "PERMANOVA (Bray-Curtis)",
    "adonis2 (999 perm.)",
    sprintf("F = %.2f", perm_bc$F[1]),
    sprintf("%.4f", perm_bc$`Pr(>F)`[1]),
    sprintf("R2 = %.4f", perm_bc$R2[1]),
    ifelse(perm_bc$`Pr(>F)`[1] > 0.05, "Not significant", "Significant"),

  "PERMANOVA (Aitchison)",
    "adonis2 (999 perm.)",
    sprintf("F = %.2f", perm_ait$F[1]),
    sprintf("%.4f", perm_ait$`Pr(>F)`[1]),
    sprintf("R2 = %.4f", perm_ait$R2[1]),
    ifelse(perm_ait$`Pr(>F)`[1] > 0.05, "Not significant", "Significant"),

  "PERMDISP (Bray-Curtis)",
    "betadisper (999 perm.)",
    sprintf("F = %.2f", pd_bc$tab$F[1]),
    sprintf("%.4f", pd_bc$tab$`Pr(>F)`[1]),
    "-",
    ifelse(pd_bc$tab$`Pr(>F)`[1] > 0.05, "Not significant", "Significant"),

  "PERMDISP (Aitchison)",
    "betadisper (999 perm.)",
    sprintf("F = %.2f", pd_ait$tab$F[1]),
    sprintf("%.4f", pd_ait$tab$`Pr(>F)`[1]),
    "-",
    ifelse(pd_ait$tab$`Pr(>F)`[1] > 0.05, "Not significant", "Significant"),

  "Diff. Abundance",
    "ALDEx2 (BH correction)",
    sprintf("%d of %d taxa", nrow(sig_aldex), nrow(aldex_res)),
    "BH-corrected",
    "-",
    ifelse(nrow(sig_aldex) == 0,
           "None significant",
           paste(nrow(sig_aldex), "significant"))
)

kable(
  summary_table,
  col.names = c("Analysis", "Method", "Test statistic", "p-value",
                "Effect size", "Conclusion"),
  align     = c("l", "l", "c", "c", "c", "l"),
  booktabs  = TRUE,
  longtable = FALSE,
  caption   = "Consolidated statistical results. All tests compare FIT vs fresh collection. BH = Benjamini--Hochberg FDR correction."
) %>%
  kable_styling(
    latex_options = c("scale_down"),
    font_size     = 9,
    full_width    = TRUE
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "2.8cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "1.2cm") %>%
  column_spec(5, width = "2cm") %>%
  column_spec(6, width = "2.2cm") %>%
  pack_rows("Alpha Diversity", 1, 2) %>%
  pack_rows("Beta Diversity -- Composition", 3, 4) %>%
  pack_rows("Beta Diversity -- Dispersion", 5, 6) %>%
  pack_rows("Taxon-Level", 7, 7)
```

```{r export-summary, include=FALSE}
write_tsv(summary_table, "NN_results_summary.tsv")
```

---

# Conclusion

This analysis applied MetaPhlAn4 species-level profiling to compare FIT and fresh collection methods using No Neoplasia (NN) samples from the Piccinno *et al.* (2025) Nature Medicine cohort. The central question was whether FIT-based collection introduces systematic bias relative to fresh faecal sampling, supporting its use for genomic biomarker identification.

Alpha diversity (Shannon, Simpson) did not differ significantly between FIT and fresh samples (Wilcoxon rank-sum, p > 0.05), with negligible effect sizes (Cliff's d approximately 0), indicating that within-sample richness and evenness are comparable across collection methods. PCoA ordination showed overlapping community structure across methods on both Bray--Curtis and Aitchison distances, with the dominant source of variation expected to be inter-individual and geographic differences rather than collection method.

PERMANOVA identified a statistically significant effect of collection method label (Bray--Curtis: F = `r round(perm_bc$F[1], 2)`, R2 = `r round(perm_bc$R2[1], 3)`, p = `r perm_bc[["Pr(>F)"]][1]`; Aitchison: F = `r round(perm_ait$F[1], 2)`, R2 = `r round(perm_ait$R2[1], 3)`, p = `r perm_ait[["Pr(>F)"]][1]`). However, because the two groups originate from different individuals on different continents, a portion of the R2 attributed to collection method may reflect geographic, dietary, or host-genetic differences between cohorts rather than a genuine collection-method effect. This interpretation is reinforced by the small absolute R2 values. PERMDISP confirmed homogeneous within-group dispersion for both distance metrics (p > 0.05).

`r if (nrow(sig_aldex) == 0) "ALDEx2 identified no differentially abundant taxa after Benjamini--Hochberg correction, providing strong evidence against collection-method-driven taxon-level bias." else paste0("ALDEx2 identified ", nrow(sig_aldex), " differentially abundant taxa after Benjamini--Hochberg correction. Given the cross-cohort design, these should be interpreted cautiously as they may reflect population-level differences rather than a collection-method effect.")` Taken together, these MetaPhlAn4-based results support the use of FIT-based collection as a method that preserves faecal microbiome composition for shotgun metagenomic analysis in the No Neoplasia context.

---

# Session Information

```{r session-info}
si <- sessionInfo()

si_df <- data.frame(
  Item  = c("R version", "Platform", "OS"),
  Value = c(si$R.version$version.string, si$platform, si$running),
  stringsAsFactors = FALSE
)

kable(si_df, col.names = c("Item", "Details"), booktabs = TRUE,
      caption = "R session information.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9)

pkg_df <- data.frame(
  Package = names(si$otherPkgs),
  Version = sapply(si$otherPkgs, function(x) x$Version),
  stringsAsFactors = FALSE
) %>% arrange(Package)

kable(pkg_df, col.names = c("Package", "Version"), booktabs = TRUE,
      caption = "Key packages loaded.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9,
                full_width = FALSE)
```
