---
title: "FIT vs Direct Collection: Paired Sample Analysis"
subtitle: "Per-subject comparison of MetaPhlAn4 metagenomic profiles"
author: "Brad Hart"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
    keep_tex: false
geometry: "margin=2cm"
fontsize: 10pt
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{colortbl}
  - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos  = "H",
  dpi     = 150
)
```

```{r load-packages, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(phyloseq)
  library(vegan)
  library(patchwork)
  library(knitr)
  library(kableExtra)
  library(RColorBrewer)
})
set.seed(42)
```

```{r theme-setup, include=FALSE}
theme_pub <- theme_bw(base_size = 10) +
  theme(
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "grey95", colour = NA),
    plot.title        = element_text(face = "bold", size = 11),
    axis.title        = element_text(size = 9),
    legend.title      = element_text(size = 8),
    legend.text       = element_text(size = 7)
  )

group_cols   <- c("direct" = "#4E79A7", "FIT" = "#E15759")
phylum_order <- c("Firmicutes", "Bacteroidota", "Actinomycetota",
                  "Pseudomonadota", "Verrucomicrobiota", "Other")
```

```{r data-load, include=FALSE}
# ── Read abundance table ──────────────────────────────────────────────────────
abund_raw <- read_tsv(
  "2020_comparison_metaphlan4_merged_FITvsNoSolution.tsv",
  show_col_types = FALSE
)

# Locate taxonomy column
clade_col <- names(abund_raw)[sapply(names(abund_raw), function(col) {
  any(str_detect(as.character(abund_raw[[col]]), fixed("k__")), na.rm = TRUE)
})]
if (clade_col != "clade_name") abund_raw <- rename(abund_raw, clade_name = all_of(clade_col))

# Filter to species level
abund_filt <- abund_raw %>%
  filter(str_detect(clade_name, "s__"), !str_detect(clade_name, "t__"))

# Parse taxonomy
tax_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
parse_metaphlan_taxonomy <- function(clade_names) {
  prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__")
  tax_df   <- matrix(NA_character_, nrow = length(clade_names), ncol = 7,
                     dimnames = list(NULL, tax_ranks))
  for (i in seq_along(prefixes)) {
    tax_df[, i] <- str_extract(clade_names, paste0(prefixes[i], "[^|]+")) %>%
      str_remove(fixed(prefixes[i]))
  }
  as.data.frame(tax_df, stringsAsFactors = FALSE)
}

tax_table_df <- parse_metaphlan_taxonomy(abund_filt$clade_name)

# Build species-level rownames
taxon_labels <- tax_table_df$Species
taxon_labels[is.na(taxon_labels)] <- tax_table_df$Genus[is.na(taxon_labels)]
taxon_labels[is.na(taxon_labels)] <- paste0("unclassified_", seq_along(which(is.na(taxon_labels))))
taxon_labels  <- str_replace_all(taxon_labels, "_", " ")
rownames(tax_table_df) <- make.unique(taxon_labels)

# OTU matrix
select   <- dplyr::select
otu_mat  <- abund_filt %>% select(-clade_name) %>% as.data.frame()
rownames(otu_mat) <- rownames(tax_table_df)

# Metadata
meta <- read_tsv("metadata.tsv", show_col_types = FALSE) %>%
  column_to_rownames("sample_id")
meta$source <- factor(meta$source, levels = c("direct", "FIT"))

shared_samples <- intersect(colnames(otu_mat), rownames(meta))
otu_mat <- otu_mat[, shared_samples]
meta    <- meta[shared_samples, , drop = FALSE]

# Build phyloseq
ps <- phyloseq(
  otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_table_df)),
  sample_data(meta)
)

# Prevalence filter (10%)
taxa_prev <- apply(otu_table(ps), 1, function(x) sum(x > 0))
ps_filt   <- prune_taxa(names(taxa_prev[taxa_prev >= ceiling(0.10 * nsamples(ps))]), ps)

# Relative abundance and CLR
ps_rel     <- transform_sample_counts(ps_filt, function(x) x / sum(x) * 100)
otu_pseudo <- as(otu_table(ps_filt), "matrix") + 0.5
clr_mat    <- t(apply(otu_pseudo, 2, function(x) log(x) - mean(log(x))))

# Subject pairing table
subjects <- unique(meta$subject_id)
pair_map <- tibble(
  subject_id = subjects,
  direct_id  = rownames(meta)[meta$subject_id == subjects & meta$source == "direct"][
    match(subjects, meta$subject_id[meta$source == "direct"])],
  fit_id     = rownames(meta)[meta$subject_id == subjects & meta$source == "FIT"][
    match(subjects, meta$subject_id[meta$source == "FIT"])]
) %>% arrange(as.numeric(str_extract(subject_id, "\\d+")))

# Rebuild pair_map properly
pair_map <- tibble(subject_id = subjects) %>%
  mutate(
    direct_id = map_chr(subject_id, ~rownames(meta)[meta$subject_id == .x & meta$source == "direct"][1]),
    fit_id    = map_chr(subject_id, ~rownames(meta)[meta$subject_id == .x & meta$source == "FIT"][1])
  ) %>%
  arrange(as.numeric(str_extract(subject_id, "\\d+")))

cat("Pairs identified:", nrow(pair_map), "\n")
```

---

# Overview

This document provides a per-subject paired analysis of MetaPhlAn4 metagenomic profiles, comparing each individual's **FIT** and **direct (No Solution)** faecal samples from the Byrd et al. (2020) dataset (BioProject PRJNA606198). Each section covers one of the `r nrow(pair_map)` subject pairs and includes:

- **Alpha diversity:** Shannon and Simpson indices, observed species richness
- **Pairwise distances:** Bray-Curtis and Aitchison (Euclidean CLR) distances between the two samples
- **Bray-Curtis similarity:** expressed as percentage similarity (1 - distance)
- **Phylum-level composition:** side-by-side bar chart
- **Bacteroidota detail:** within-Bacteroidota genus breakdown
- **Firmicutes detail:** within-Firmicutes genus breakdown
- **Top species:** ranked table showing direct vs FIT relative abundance and absolute difference

All relative abundances are expressed as percentage of the whole community unless otherwise noted. Within-phylum figures rescale to 100% within that phylum to isolate compositional shifts independently of global phylum abundance.

A **cross-pair summary table** at the end collates key metrics across all subjects.

---

```{r pair-functions, include=FALSE}
# ── Helper: alpha diversity for a single sample ───────────────────────────────
sample_alpha <- function(sample_id) {
  counts <- round(as.numeric(otu_table(ps_filt)[, sample_id]) / 
                    sum(otu_table(ps_filt)[, sample_id]) * 10000)
  ab     <- counts[counts > 0]
  n      <- sum(ab)
  H      <- -sum((ab / n) * log(ab / n))       # Shannon
  D      <- 1 - sum((ab / n)^2)                # Simpson (1-D)
  S      <- sum(counts > 0)                    # Observed species
  c(Shannon = round(H, 3), Simpson = round(D, 4), Richness = S)
}

# ── Helper: pairwise distances ─────────────────────────────────────────────────
pair_distances <- function(direct_id, fit_id) {
  # Bray-Curtis on relative abundance
  rel_mat <- t(as(otu_table(ps_rel), "matrix"))
  bc      <- as.numeric(vegdist(rel_mat[c(direct_id, fit_id), ], method = "bray"))
  # Aitchison (Euclidean CLR)
  ait     <- as.numeric(dist(clr_mat[c(direct_id, fit_id), ], method = "euclidean"))
  c(BrayCurtis = round(bc, 4),
    Similarity  = round((1 - bc) * 100, 1),
    Aitchison   = round(ait, 2))
}

# ── Helper: phylum bar plot for a pair ────────────────────────────────────────
plot_phylum_pair <- function(direct_id, fit_id, subject) {
  ps_sub <- prune_samples(c(direct_id, fit_id), ps_rel)
  ps_phy <- tax_glom(ps_sub, taxrank = "Phylum", NArm = FALSE)

  df <- psmelt(ps_phy) %>%
    mutate(
      Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum),
      Source = factor(source, levels = c("direct", "FIT"),
                      labels = c("Direct", "FIT"))
    )

  # head() after slice_max prevents tie-expansion from inflating the list
  n_top   <- 7
  top_phy <- df %>%
    group_by(Phylum) %>% summarise(m = mean(Abundance)) %>%
    arrange(desc(m)) %>%
    head(n_top) %>%
    pull(Phylum)

  df <- df %>% mutate(Phylum = ifelse(Phylum %in% top_phy, Phylum, "Other"))

  n_cols <- max(3, length(top_phy))  # brewer.pal minimum is 3
  pal <- c(setNames(brewer.pal(n_cols, "Set2")[seq_along(top_phy)],
                    sort(top_phy)),
           "Other" = "grey70")

  ggplot(df, aes(x = Source, y = Abundance, fill = Phylum)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_fill_manual(values = pal) +
    scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
    labs(title = paste0(subject, ": Phylum Composition"),
         x = NULL, y = "Relative abundance (%)") +
    theme_pub +
    theme(legend.position = "right", legend.text = element_text(size = 7))
}

# ── Helper: within-phylum genus bar for a pair ────────────────────────────────
plot_within_phylum <- function(direct_id, fit_id, subject, phylum,
                                fill_high = "#4E79A7") {
  ps_sub  <- prune_samples(c(direct_id, fit_id), ps_rel)

  # subset_taxa() evaluates in the tax_table environment and cannot see
  # function arguments as variables. Use prune_taxa() with an explicit
  # logical vector instead.
  phy_vec <- as.character(tax_table(ps_sub)[, "Phylum"])
  keep    <- taxa_names(ps_sub)[!is.na(phy_vec) & phy_vec == phylum]
  if (length(keep) == 0) return(NULL)
  ps_phy  <- prune_taxa(keep, ps_sub)

  ps_genus <- tax_glom(ps_phy, taxrank = "Genus", NArm = TRUE)
  ps_resc  <- transform_sample_counts(ps_genus, function(x) x / sum(x) * 100)

  df <- psmelt(ps_resc) %>%
    mutate(
      Genus  = ifelse(is.na(Genus) | Genus == "", "Unclassified", Genus),
      Source = factor(source, levels = c("direct", "FIT"),
                      labels = c("Direct", "FIT"))
    )

  # head() clamps to exactly n_top after arrange to avoid tie expansion
  n_top <- 8
  top_g <- df %>%
    group_by(Genus) %>% summarise(m = mean(Abundance)) %>%
    arrange(desc(m)) %>%
    head(n_top) %>%
    pull(Genus)

  df <- df %>% mutate(Genus = ifelse(Genus %in% top_g, Genus, "Other"))

  n_cols <- max(3, length(top_g))
  pal <- c(setNames(brewer.pal(n_cols, "Paired")[seq_along(top_g)],
                    sort(top_g)),
           "Other" = "grey70")

  ggplot(df, aes(x = Source, y = Abundance, fill = Genus)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_fill_manual(values = pal) +
    scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
    labs(title = paste0(phylum, " (within-phylum)"),
         x = NULL, y = "Within-phylum abundance (%)") +
    theme_pub +
    theme(legend.position = "right",
          legend.text = element_text(face = "italic", size = 7))
}

# ── Helper: paired alpha diversity dot plot for one subject ───────────────────
plot_alpha_pair <- function(alpha_d, alpha_f, subject) {
  df <- tibble(
    Metric = rep(c("Shannon index", "Simpson index (1-D)", "Observed species"), 2),
    Method = rep(c("Direct", "FIT"), each = 3),
    Value  = c(alpha_d["Shannon"], alpha_d["Simpson"], alpha_d["Richness"],
               alpha_f["Shannon"], alpha_f["Simpson"], alpha_f["Richness"])
  ) %>%
    mutate(
      Method = factor(Method, levels = c("Direct", "FIT")),
      Metric = factor(Metric, levels = c("Shannon index",
                                          "Simpson index (1-D)",
                                          "Observed species"))
    )

  # Connecting line data
  line_df <- df %>%
    select(Metric, Method, Value) %>%
    pivot_wider(names_from = Method, values_from = Value) %>%
    mutate(colour = ifelse(FIT >= Direct, "#E15759", "#4E79A7"))

  ggplot() +
    # Connecting line between the two points
    geom_segment(data = line_df,
                 aes(x = "Direct", xend = "FIT",
                     y = Direct,   yend = FIT,
                     colour = colour),
                 linewidth = 0.7, alpha = 0.6) +
    # Points
    geom_point(data = df,
               aes(x = Method, y = Value, colour = Method),
               size = 4) +
    scale_colour_manual(values = c("Direct"   = "#4E79A7",
                                   "FIT"      = "#E15759",
                                   "#4E79A7"  = "#4E79A7",
                                   "#E15759"  = "#E15759")) +
    facet_wrap(~Metric, scales = "free_y", nrow = 1) +
    labs(title   = paste0(subject, ": Alpha Diversity"),
         x = NULL, y = "Value") +
    theme_pub +
    theme(legend.position = "none",
          strip.text      = element_text(size = 8))
}

# ── Helper: top species table ─────────────────────────────────────────────────
top_species_table <- function(direct_id, fit_id, n = 20) {
  rel_mat <- as(otu_table(ps_rel), "matrix")
  df <- tibble(
    Species    = rownames(rel_mat),
    Direct     = as.numeric(rel_mat[, direct_id]),
    FIT        = as.numeric(rel_mat[, fit_id])
  ) %>%
    mutate(
      Mean  = (Direct + FIT) / 2,
      Diff  = FIT - Direct,
      AbsDiff = abs(Diff)
    ) %>%
    filter(Mean > 0) %>%
    slice_max(Mean, n = n) %>%
    arrange(desc(Mean)) %>%
    mutate(across(c(Direct, FIT, Mean, Diff), ~round(.x, 3)))

  df
}
```

```{r generate-pairs, results='asis', fig.width=7, fig.height=3.5}
# Store cross-pair summary data
summary_rows <- list()

for (i in seq_len(nrow(pair_map))) {

  subj       <- pair_map$subject_id[i]
  direct_id  <- pair_map$direct_id[i]
  fit_id     <- pair_map$fit_id[i]
  subj_label <- paste0("Subject ", str_extract(subj, "\\d+"))

  # ── Section header ────────────────────────────────────────────────────────
  cat("\n\n---\n\n")
  cat("# ", subj_label, " (", direct_id, " / ", fit_id, ")\n\n", sep = "")

  # ── Alpha diversity ───────────────────────────────────────────────────────
  alpha_d <- sample_alpha(direct_id)
  alpha_f <- sample_alpha(fit_id)

  alpha_tbl <- tibble(
    Metric   = c("Shannon index", "Simpson index (1-D)", "Observed species"),
    Direct   = as.character(c(alpha_d["Shannon"], alpha_d["Simpson"],
                               alpha_d["Richness"])),
    FIT      = as.character(c(alpha_f["Shannon"], alpha_f["Simpson"],
                               alpha_f["Richness"])),
    Delta    = c(
      round(alpha_f["Shannon"]  - alpha_d["Shannon"],  3),
      round(alpha_f["Simpson"]  - alpha_d["Simpson"],  4),
      alpha_f["Richness"] - alpha_d["Richness"]
    )
  )

  cat("## Alpha Diversity\n\n")
  print(kable(alpha_tbl,
              col.names = c("Metric", "Direct", "FIT", "Delta (FIT - Direct)"),
              booktabs = TRUE, align = c("l","c","c","c"),
              caption = paste0(subj_label, ": Alpha diversity comparison.")) %>%
    kable_styling(latex_options = "hold_position", font_size = 9) %>%
    row_spec(which(abs(as.numeric(alpha_tbl$Delta)) > 0),
             background = "#f7f7f7"))
  cat("\n\n")

  # Paired dot plot: one point per method, line connecting the pair
  print(plot_alpha_pair(alpha_d, alpha_f, subj_label))
  cat("\n\n")

  # ── Pairwise distances ────────────────────────────────────────────────────
  dists <- pair_distances(direct_id, fit_id)

  dist_tbl <- tibble(
    Metric    = c("Bray-Curtis distance", "Bray-Curtis similarity (%)",
                  "Aitchison distance"),
    Value     = as.character(dists)
  )

  cat("## Pairwise Distances\n\n")
  print(kable(dist_tbl,
              col.names = c("Metric", "Value"),
              booktabs = TRUE, align = c("l","c"),
              caption = paste0(subj_label, ": Pairwise distance between direct and FIT samples.")) %>%
    kable_styling(latex_options = "hold_position", font_size = 9,
                  full_width = FALSE))
  cat("\n\n")

  # ── Phylum + within-phylum plots ──────────────────────────────────────────
  cat("## Community Composition\n\n")

  p1 <- plot_phylum_pair(direct_id, fit_id, subj_label)
  p2 <- plot_within_phylum(direct_id, fit_id, subj_label,
                            "Bacteroidota", "#4E79A7")
  p3 <- plot_within_phylum(direct_id, fit_id, subj_label,
                            "Firmicutes",   "#E15759")

  # Phylum overview
  print(p1)
  cat("\n\n")

  # Bacteroidota + Firmicutes side by side if both exist
  if (!is.null(p2) && !is.null(p3)) {
    combined <- p2 + p3 +
      plot_annotation(
        title = paste0(subj_label, ": Within-Phylum Genus Composition"),
        theme = theme(plot.title = element_text(face = "bold", size = 11))
      ) +
      plot_layout(guides = "collect") &
      theme(legend.position = "right")
    print(combined)
  } else if (!is.null(p2)) {
    print(p2)
  } else if (!is.null(p3)) {
    print(p3)
  }
  cat("\n\n")

  # ── Top species table ─────────────────────────────────────────────────────
  cat("## Top 20 Species by Mean Abundance\n\n")

  sp_tbl <- top_species_table(direct_id, fit_id, n = 20)

  # Colour coding: positive diff = higher in FIT (red), negative = higher in direct (blue)
  sp_display <- sp_tbl %>%
    select(Species, Direct, FIT, Diff) %>%
    mutate(Direction = case_when(
      Diff >  0.5 ~ "Higher in FIT",
      Diff < -0.5 ~ "Higher in Direct",
      TRUE        ~ "Similar"
    ))

  higher_fit_rows    <- which(sp_display$Direction == "Higher in FIT")
  higher_direct_rows <- which(sp_display$Direction == "Higher in Direct")

  k_out <- kable(
    sp_display %>% select(-Direction),
    col.names = c("Species", "Direct (%)", "FIT (%)", "Diff (FIT - Direct)"),
    booktabs  = TRUE,
    align     = c("l","c","c","c"),
    caption   = paste0(subj_label,
      ": Top 20 species by mean relative abundance. ",
      "Blue shading = higher in direct (diff < -0.5); ",
      "Red shading = higher in FIT (diff > +0.5).")
  ) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"),
                  font_size = 8) %>%
    column_spec(1, italic = TRUE, width = "5cm")

  if (length(higher_direct_rows) > 0)
    k_out <- row_spec(k_out, higher_direct_rows, background = "#dce9f5")
  if (length(higher_fit_rows) > 0)
    k_out <- row_spec(k_out, higher_fit_rows,    background = "#fde3e3")

  print(k_out)
  cat("\n\n")

  # ── Species unique to each method for this subject ────────────────────────
  cat("## Species Detected in Only One Method\n\n")

  rel_mat    <- as(otu_table(ps_rel), "matrix")
  only_direct <- rownames(rel_mat)[rel_mat[, direct_id] > 0 &
                                    rel_mat[, fit_id]    == 0]
  only_fit    <- rownames(rel_mat)[rel_mat[, direct_id] == 0 &
                                    rel_mat[, fit_id]    > 0]

  unique_tbl <- bind_rows(
    tibble(Method = "Direct only",
           Species = only_direct,
           Abundance = round(rel_mat[only_direct, direct_id], 3)),
    tibble(Method = "FIT only",
           Species = only_fit,
           Abundance = round(rel_mat[only_fit, fit_id], 3))
  ) %>%
    arrange(Method, desc(Abundance))

  if (nrow(unique_tbl) == 0) {
    cat("No species detected exclusively in one method for this subject.\n\n")
  } else {
    print(kable(unique_tbl,
                col.names = c("Detected in", "Species", "Relative abundance (%)"),
                booktabs = TRUE, align = c("l","l","c"),
                caption = paste0(subj_label,
                  ": Species detected in only one collection method.")) %>%
      kable_styling(latex_options = c("hold_position", "scale_down"),
                    font_size = 8) %>%
      column_spec(2, italic = TRUE, width = "6cm"))
    cat("\n\n")
  }

  # ── Store summary row ─────────────────────────────────────────────────────
  summary_rows[[i]] <- tibble(
    Subject         = subj_label,
    Direct_ID       = direct_id,
    FIT_ID          = fit_id,
    Shannon_Direct  = alpha_d["Shannon"],
    Shannon_FIT     = alpha_f["Shannon"],
    Shannon_Delta   = round(alpha_f["Shannon"] - alpha_d["Shannon"], 3),
    Simpson_Direct  = alpha_d["Simpson"],
    Simpson_FIT     = alpha_f["Simpson"],
    Simpson_Delta   = round(alpha_f["Simpson"] - alpha_d["Simpson"], 4),
    Richness_Direct = alpha_d["Richness"],
    Richness_FIT    = alpha_f["Richness"],
    Richness_Delta  = alpha_f["Richness"] - alpha_d["Richness"],
    BC_Distance     = dists["BrayCurtis"],
    BC_Similarity   = dists["Similarity"],
    Aitchison_Dist  = dists["Aitchison"],
    N_Only_Direct   = length(only_direct),
    N_Only_FIT      = length(only_fit)
  )
}
```

```{r build-summary-df, include=FALSE}
# Build summary_df here so all subsequent chunks can access it
summary_df <- bind_rows(summary_rows)
```

---

# Cross-Pair Summary

Key metrics across all `r nrow(pair_map)` subject pairs are collated
below. For legibility the summary is split into two tables: alpha
diversity per subject (Table \@ref(tab:cross-pair-alpha)) and pairwise
distances plus method-unique species counts (Table
\@ref(tab:cross-pair-beta)).

```{r cross-pair-alpha}
kable(
  summary_df %>%
    select(Subject, Shannon_Direct, Shannon_FIT, Shannon_Delta,
           Simpson_Direct, Simpson_FIT, Simpson_Delta,
           Richness_Direct, Richness_FIT, Richness_Delta),
  col.names = c("Subject",
                "Direct", "FIT", "Delta",
                "Direct", "FIT", "Delta",
                "Direct", "FIT", "Delta"),
  booktabs = TRUE,
  align    = c("l", rep("c", 9)),
  caption  = "Cross-pair alpha-diversity summary. Delta = FIT - Direct.
              Red text indicates higher in FIT; blue indicates higher in Direct."
) %>%
  kable_styling(latex_options = "hold_position", font_size = 8) %>%
  add_header_above(c(" " = 1, "Shannon" = 3,
                     "Simpson (1-D)" = 3, "Observed richness" = 3),
                   bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(4,  color = ifelse(summary_df$Shannon_Delta  > 0, "#c0392b", "#2980b9")) %>%
  column_spec(7,  color = ifelse(summary_df$Simpson_Delta  > 0, "#c0392b", "#2980b9")) %>%
  column_spec(10, color = ifelse(summary_df$Richness_Delta > 0, "#c0392b", "#2980b9"))
```

```{r cross-pair-beta}
kable(
  summary_df %>%
    select(Subject, BC_Distance, BC_Similarity, Aitchison_Dist,
           N_Only_Direct, N_Only_FIT),
  col.names = c("Subject",
                "Bray-Curtis distance",
                "Bray-Curtis similarity (%)",
                "Aitchison distance",
                "Direct-only species",
                "FIT-only species"),
  booktabs = TRUE,
  align    = c("l", rep("c", 5)),
  caption  = "Cross-pair beta-diversity distances (Bray-Curtis and Aitchison)
              and number of species detected exclusively in one collection method."
) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  row_spec(0, bold = TRUE)
```

```{r alpha-boxplots, fig.width=7, fig.height=4.5, out.width="\\linewidth", fig.cap="Group-level box plots for alpha diversity metrics across all 12 subject pairs. Each point represents one subject; lines connect paired direct and FIT samples from the same individual. Blue = direct, red = FIT."}

# Long format for box plots with pairing info
alpha_box_df <- summary_df %>%
  select(Subject,
         Shannon_Direct, Shannon_FIT,
         Simpson_Direct, Simpson_FIT,
         Richness_Direct, Richness_FIT) %>%
  pivot_longer(-Subject,
               names_to  = c("Metric", "Method"),
               names_sep = "_",
               values_to = "Value") %>%
  mutate(
    Method = factor(Method, levels = c("Direct", "FIT")),
    Metric = factor(Metric,
                    levels = c("Shannon", "Simpson", "Richness"),
                    labels = c("Shannon index",
                               "Simpson index (1-D)",
                               "Observed species"))
  )

ggplot(alpha_box_df, aes(x = Method, y = Value, fill = Method)) +
  geom_boxplot(width = 0.45, alpha = 0.55, outlier.shape = NA) +
  # Lines connecting paired samples
  geom_line(aes(group = Subject), colour = "grey60",
            linewidth = 0.4, alpha = 0.7) +
  geom_point(aes(colour = Method), size = 2.5, alpha = 0.9) +
  scale_fill_manual(values  = group_cols) +
  scale_colour_manual(values = group_cols) +
  facet_wrap(~Metric, scales = "free_y", nrow = 1) +
  labs(
    title    = "Alpha Diversity: All Subject Pairs (Direct vs FIT)",
    subtitle = "Lines connect paired samples from the same individual",
    x = NULL, y = "Value"
  ) +
  theme_pub +
  theme(legend.position = "none",
        strip.text      = element_text(size = 9))
```

```{r summary-plots, fig.width=7, fig.height=8, out.width="\\linewidth", fig.cap="Cross-pair summary plots. Top row: Shannon index and species richness per subject, coloured by method. Middle row: Bray-Curtis similarity and Aitchison distance per pair. Bottom row: species detected uniquely in each method per subject."}

# Reshape for per-subject bar plots
summary_long_alpha <- summary_df %>%
  select(Subject, Shannon_Direct, Shannon_FIT, Richness_Direct, Richness_FIT) %>%
  pivot_longer(-Subject,
               names_to  = c("Metric", "Method"),
               names_sep = "_",
               values_to = "Value") %>%
  mutate(Method = factor(Method, levels = c("Direct", "FIT")))

p_shannon_cross <- summary_long_alpha %>%
  filter(Metric == "Shannon") %>%
  ggplot(aes(x = Subject, y = Value, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = group_cols) +
  labs(title = "Shannon Index", x = NULL, y = "Shannon index") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        legend.position = "none")

p_richness_cross <- summary_long_alpha %>%
  filter(Metric == "Richness") %>%
  ggplot(aes(x = Subject, y = Value, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = group_cols) +
  labs(title = "Species Richness", x = NULL, y = "Observed species") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p_bc_cross <- summary_df %>%
  ggplot(aes(x = Subject, y = BC_Similarity)) +
  geom_col(fill = "#7fb3d3", width = 0.6) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey40") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Bray-Curtis Similarity (%)",
       subtitle = "Dashed = 50% threshold",
       x = NULL, y = "Similarity (%)") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p_ait_cross <- summary_df %>%
  ggplot(aes(x = Subject, y = Aitchison_Dist)) +
  geom_col(fill = "#f0a58b", width = 0.6) +
  labs(title = "Aitchison Distance",
       x = NULL, y = "Euclidean CLR distance") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p_unique_cross <- summary_df %>%
  select(Subject, N_Only_Direct, N_Only_FIT) %>%
  pivot_longer(-Subject, names_to = "Method", values_to = "N") %>%
  mutate(Method = recode(Method,
                          N_Only_Direct = "Direct only",
                          N_Only_FIT    = "FIT only")) %>%
  ggplot(aes(x = Subject, y = N, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Direct only" = "#4E79A7",
                                "FIT only"    = "#E15759")) +
  labs(title = "Species Detected in Only One Method",
       x = NULL, y = "Number of species") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

(p_shannon_cross + p_richness_cross) /
  (p_bc_cross     + p_ait_cross) /
  (p_unique_cross + plot_spacer()) +
  plot_annotation(
    title = "Cross-Pair Summary",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )
```

```{r export-summary, include=FALSE}
write_tsv(summary_df, "paired_analysis_summary.tsv")
```

---

# Variance Decomposition: Collection Method vs Inter-Individual Variation

A core question for the FIT versus direct comparison is how much of the
total compositional variance across these 24 samples is attributable to
the collection method itself, and how much reflects genuine
inter-individual differences between subjects. Two complementary
analyses address this:

- **Beta-diversity variance** is partitioned by PERMANOVA
  (`vegan::adonis2`) on the full pairwise distance matrix, fitting
  `subject_id + source` so that subject absorbs all inter-individual
  variance and `source` captures the within-subject method effect.
  Both Bray-Curtis and Aitchison (Euclidean CLR) distances are reported.
- **Alpha-diversity variance** is partitioned by balanced two-way ANOVA
  on Shannon, Simpson, and observed richness. With one observation per
  Subject:Method cell the residual term corresponds to the
  Subject:Method interaction, i.e. the extent to which the method effect
  differs across individuals.

```{r variance-beta}
# ── Build distance matrices over all 24 samples ───────────────────────────
rel_mat_all <- t(as(otu_table(ps_rel), "matrix"))
bc_all      <- vegan::vegdist(rel_mat_all, method = "bray")
ait_all     <- dist(clr_mat, method = "euclidean")

# Align metadata to distance-matrix sample order
meta_perm <- meta %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::filter(sample_id %in% labels(bc_all)) %>%
  dplyr::arrange(match(sample_id, labels(bc_all)))

# PERMANOVA — sequential SS with subject first, then source
set.seed(42)
perm_bc  <- vegan::adonis2(bc_all  ~ subject_id + source,
                           data = meta_perm,
                           by = "terms", permutations = 999)
set.seed(42)
perm_ait <- vegan::adonis2(ait_all ~ subject_id + source,
                           data = meta_perm,
                           by = "terms", permutations = 999)

build_perm_row <- function(perm_res, dist_name) {
  r <- as.data.frame(perm_res)
  tibble(
    Distance            = dist_name,
    Inter_individual_R2 = round(r["subject_id", "R2"] * 100, 1),
    Method_R2           = round(r["source",     "R2"] * 100, 1),
    Residual_R2         = round(r["Residual",   "R2"] * 100, 1),
    Method_p            = format.pval(r["source", "Pr(>F)"], digits = 2, eps = 1e-3)
  )
}

perm_summary <- bind_rows(
  build_perm_row(perm_bc,  "Bray-Curtis"),
  build_perm_row(perm_ait, "Aitchison (Euclidean CLR)")
)

kable(perm_summary,
      col.names = c("Distance metric",
                    "Inter-individual (%)",
                    "Collection method (%)",
                    "Residual (%)",
                    "Method p-value"),
      booktabs = TRUE, align = c("l", "c", "c", "c", "c"),
      caption  = "Beta-diversity variance partitioning by PERMANOVA
                  (adonis2, 999 permutations, sequential SS with subject
                  fitted first). R-squared values give the proportion of
                  total variance attributable to each term.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  row_spec(0, bold = TRUE)
```

```{r variance-alpha}
# ── Alpha diversity: balanced two-way ANOVA decomposition ────────────────
alpha_long_var <- summary_df %>%
  select(Subject,
         Shannon_Direct, Shannon_FIT,
         Simpson_Direct, Simpson_FIT,
         Richness_Direct, Richness_FIT) %>%
  pivot_longer(-Subject,
               names_to  = c("Metric", "Method"),
               names_sep = "_",
               values_to = "Value") %>%
  mutate(Subject = factor(Subject),
         Method  = factor(Method, levels = c("Direct", "FIT")))

calc_alpha_var <- function(metric_name) {
  d   <- alpha_long_var %>% dplyr::filter(Metric == metric_name)
  fit <- aov(Value ~ Subject + Method, data = d)
  tab <- summary(fit)[[1]]
  ss  <- tab[, "Sum Sq"]
  names(ss) <- trimws(rownames(tab))
  total <- sum(ss)
  tibble(
    Metric                = metric_name,
    Inter_individual_pct  = round(ss["Subject"]   / total * 100, 1),
    Method_pct            = round(ss["Method"]    / total * 100, 1),
    Residual_pct          = round(ss["Residuals"] / total * 100, 1)
  )
}

alpha_var_summary <- bind_rows(lapply(
  c("Shannon", "Simpson", "Richness"),
  calc_alpha_var
))

kable(alpha_var_summary,
      col.names = c("Alpha metric",
                    "Inter-individual (%)",
                    "Collection method (%)",
                    "Residual (%)"),
      booktabs = TRUE, align = c("l", "c", "c", "c"),
      caption  = "Alpha-diversity variance decomposition (Type I sums of
                  squares from a balanced two-way ANOVA with Subject and
                  Method main effects). The residual captures the
                  Subject-by-Method interaction, i.e. how variable the
                  method effect is across individuals.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  row_spec(0, bold = TRUE)
```

```{r variance-plot, fig.width=7, fig.height=3.8, out.width="\\linewidth", fig.cap="Stacked-bar visualisation of the variance partitioning. Each metric is normalised to 100% of total variance; inter-individual variation dominates throughout, with the collection-method effect contributing a substantially smaller share."}

var_plot_df <- bind_rows(
  perm_summary %>%
    select(Metric = Distance,
           `Inter-individual` = Inter_individual_R2,
           `Collection method` = Method_R2,
           Residual = Residual_R2) %>%
    mutate(Group = "Beta diversity"),
  alpha_var_summary %>%
    select(Metric,
           `Inter-individual` = Inter_individual_pct,
           `Collection method` = Method_pct,
           Residual = Residual_pct) %>%
    mutate(Group = "Alpha diversity")
) %>%
  pivot_longer(c(`Inter-individual`, `Collection method`, Residual),
               names_to = "Source", values_to = "Percent") %>%
  mutate(
    Source = factor(Source,
                    levels = c("Inter-individual", "Collection method", "Residual")),
    Metric = factor(Metric, levels = c("Shannon", "Simpson", "Richness",
                                       "Bray-Curtis", "Aitchison (Euclidean CLR)"))
  )

ggplot(var_plot_df,
       aes(x = Metric, y = Percent, fill = Source)) +
  geom_bar(stat = "identity", width = 0.65) +
  facet_wrap(~Group, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = c("Inter-individual"  = "#4E79A7",
                                "Collection method" = "#E15759",
                                "Residual"          = "grey70")) +
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
  labs(title = "Variance Partitioning: Inter-Individual vs Collection Method",
       x = NULL, y = "Percentage of total variance (%)",
       fill = NULL) +
  theme_pub +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1, size = 8),
        legend.position  = "bottom",
        strip.text       = element_text(face = "bold"))
```

```{r export-variance, include=FALSE}
write_tsv(perm_summary,      "variance_decomposition_beta.tsv")
write_tsv(alpha_var_summary, "variance_decomposition_alpha.tsv")
```

---

# Session Information

```{r session-info}
si <- sessionInfo()

si_df <- data.frame(
  Item  = c("R version", "Platform", "OS"),
  Value = c(si$R.version$version.string, si$platform, si$running),
  stringsAsFactors = FALSE
)

kable(si_df, col.names = c("Item", "Details"), booktabs = TRUE,
      caption = "R session information.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9)
```
